KIN: Method to infer relatedness using low coverage ancient DNA

### 2022-03-31 paper_overleaf/main.tex

\title{KIN: Method to infer relatedness using low coverage ancient DNA}
\author{Divya Ratan Popli, Benjamin M. Peter}
\date{5 August 2021}
\linenumbers

@@

\noindent Knowledge of familial relationships is an important part of several research fields. However, estimating relatedness from ancient DNA can be difficult since DNA sequences extracted from ancient remains may have low coverage, ascertainment bias (if the sequences have been captured with DNA probes), and contamination from modern human DNA. In addition, the population of interest may have long runs of homozygosity (ROH) due to recent inbreeding or/and small population size. This can affect the relatedness estimates. Here, we present a Hidden Markov Model (HMM) to estimate identity by descent (IBD) fragments and degree of relatedness up to 3rd degree, while additionally differentiating between siblings and parent-child. The output can be further refined with another HMM that we developed to estimate ROH proportions.
\end{abstract}

@@

Identifying related individuals is a common task in applications of genetics. Relatedness is of direct interest in e.g. DNA forensics, where familial search can aid in solving criminal cases, and to identify missing persons following a disaster \cite{murphy_law_2018,ram_genealogy_2018}. Genetic paternity tests have an important application in resolving family relation, e.g. in establishing relationship between a person applying for immigration and the claimed relatives \cite{egeland_beyond_2000}. It is also an essential pre-processing step in population genetics and association studies, where samples are typically assumed to be independent random draws from the population.

In ancient DNA studies, relatedness can be used to identify bones and teeth belonging to the same individual, and can provide an understanding of an ancient society's organization and hierarchy, social structures, and cultural aspects ~\cite{baca_ancient_2012,mittnik_KINship-based_2019,sikora_ancient_2017}. For animal and plant breeders and conservation biologists, reconstructing pedigrees and finding related individuals is important to ensure diversity. ~\cite{habier_impact_2007,oliehoek_estimating_2006,kardos_measuring_2015}

\subsection{Simple way to get relatedness, methods that do it with a lot of data}

Commonly, pairs of related individuals are identified by looking for parts of the genome that are identical by descent (IBD), ie. inherited from a recent common ancestor. Due to the laws of Mendelian segregation, a parent, for example, will share exactly one set of chromosome IBD with its offspring, with the other set of chromosomes in the offspring coming from the other parent. A grandparent will, on average share a quarter of its genome with a grand-child, because recombination broke up the chromosomes.

However, it is not possible to directly observe IBD segments, and so most methods infer them by first identifying segments of the genome that are Identical by State (IBS) and using population allele frequencies to calculate the probability of IBD given IBS \cite{vai_KINship_2020}. One exception is SNPduo, a software that uses IBS direclty to visualize and analyze pairwise relatedness. \cite{roberson_visualization_2009}. Most methods, however, use IBD as an intermediary in inferring relatedness \cite{boehnke_accurate_1997,lynch_estimation_1999, albrechtsen_natural_2010, purcell_plink_2007}. For a non-inbred population, there are three IBD states possible: either no chromosomes are shared, only one chromosome is shared, or both chromosomes are shared. The genome-wide proportions of these states (usually referred to as $k_0$, $k_1$, $k_2$, so that $k_0+k_1+k_2=1$) can be used to infer the degree and nature of relatedness for a pair of individuals. For example, a pair of siblings are expected to have all three possible IBD states with probabilities (0.25,0.5,0.25) as shown in Fig \ref{fig0:schematic}. These IBD probabilities can directly be used to categorize their relatedness as shown in table \ref{tab:Table 1}. One can also use these probabilities to estimate the coefficient of relatedness as $r= k_1/2 + k_2$. The coefficient of relatedness (r) is defined as the proportion of genome IBD for a pair of individuals.

@@

There are numerous methods that utilize population allele frequencies, phase information, recombination maps, or genotype calls to estimate the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,li_accurate_2014,thornton_estimating_2012}. PLINK, a popular tool to infer relatedness coefficient in diploid genotype data, estimates IBD probabilities from observed IBS, using the allele frequencies at each Single Nucleotide Polymorphism (SNP) in the analyzed samples \cite{purcell_plink_2007}. KING is another widely used software that allows relatedness inference with genotype data in homogeneous populations using population allele frequencies in the samples (KING homo), but can also work with structured populations with no good population allele frequencies using heterozygosity estimates from each sample (KING robust) \cite{manichaikul_robust_2010}. .

\begin{table}
\caption{\label{tab:Table 1}IBD sharing probabilities for different relations}
\begin{tabular}{|c|c|c|c|}
    \hline
    Relatedness & $k_0$ & $k_1$ & $k_2$\\
    \hline
    Unrelated & 1 & 0 & 0\\
    \hline
    3rd Degree & 0.75 & 0.25 & 0\\
    \hline
    2nd Degree & 0.5 & 0.5 & 0\\
    \hline
    Siblings & 0.25 & 0.5 & 0.25\\
    \hline
    Parent-Child & 0 & 1 & 0\\
    \hline
    Identical/twins & 0 & 0 & 1\\
    \hline
\end{tabular}
\label{table1}
\end{table}

\subsection{Problems with ancient data, and methods that address these issues}
One major issue with applying the above mentioned methods to ancient DNA is that, frequently, the amount of endogenous DNA is very low, making it very difficult to get genotype calls.
Several methods surmount this problem by using genotype likelihoods \cite{lipatov_maximum_2015,korneliussen_ngsrelate_2015}.In this way, it is possible to account for the uncertainty in genotype calls by summing over all possible genotypes weighted by their genotype likelihoods. However, these approaches typically still require at least 2x coverage, since genotype likelihoods may not be very informative at lower coverages. Apart from the low amount of endogenous DNA sequences present, ancient DNA analyses often have additional challenges like contamination from present-day DNA \cite{peyregne_authentict_2020}, absence of a good reference panel, and ascertainment bias in case of captured sequences. In cases where the allele frequency or a reference panel for the target population is not available, it is possible to use allele frequency from a modern population from the same geographic location, or to try using population allele frequencies from multiple potentially close populations. However, incorrect assumptions may cause the individuals in the target population to look more similar or different to each other \cite{amorim_understanding_2018}. Several methods have been proposed to estimate relatedness without a reference panel, but many of these require either $>4x$ coverage \cite{waples_allele_2019}, or a large sample size to get an estimate of population allele frequencies from the samples \cite{theunert_joint_2017}. In this study, we show that contamination can cause some of the individuals in ancient populations to look more unrelated than they are. This happens because of comparison of endogenous reads to contaminating reads which yields more differences than comparison of endogenous reads from the ancient individuals. It is worth mentioning that different DNA libraries may have different levels of contamination, even when they have been sampled from the same specimen. Hence, contamination can reduce the power to detect related pairs, especially when the target population is quite diverged from contaminating population(s). In addition, the genomes of target individuals may have long runs of homozygosity (ROH) due to a small population size, or recent inbreeding. Long ROH causes related individuals to seem genetically closer to each other, while does not affect the genetic distance between unrelated individuals. This effect may cause an increase in false positives. Moreover, ancient DNA is commonly captured with a SNP array to enrich for endogenous sequences. The way that these SNPs are defined (e.g. selecting SNPs variable in certain individuals) becomes relevant, because relatedness methods based on the fraction of sites in different IBS states are highly sensitive to ascertainment bias. KING robust, for example, differentiates between different relatedness by plotting KING‐robust KINship against fraction of sites in IBS0 \cite{manichaikul_robust_2010}. Another method from Rosenberg,2006 \cite{rosenberg_standardized_2006} uses the fraction of sites in all three IBS states to classify related individuals in the HGDP-CEPH Human Genome Diversity Cell Line Panel. These methods may give different results when applied to data with different ascertainment schemes \cite{waples_allele_2019}.

@@

\section{Methods}

The core of  KIN is a HMM that aims to infer relatedness and IBD sharing between a pair of low-coverage individuals, optionally taking ROH tracts and contamination estimates in each individual into account. If the locations of ROH tracts are unknown, we provide another HMM to coarsely estimate the location of ROH for samples  of sufficient coverage ($\geq 0.5x$). Our method is available on \url{https://github.com/DivyaratanPopli/KIN_snakemake} along with a \textt{snakemake} \cite{koster_snakemakescalable_2012} pipeline to generate the input files for the models directly from bam files.

\subsection{Relatedness Model overview} 
Our relatedness model assumes we have low-coverage genome sequences for a number of individuals from the same population, and our goal is to infer which pairs of individuals are related. We subdivide the genome into $L$ large windows (typically of size 10Mb). The inputs of our algorithm are i) the number of overlapping sites for the $w$-th window $N_w$ for which both samples have at least one read available, ii) the expected number of pairwise differences $D_w$ at these sites, and iii) the probability of ROH in windows, by default obtained from ROH-HMM described in a following section. $D_w$ is the number of sites that are different between a pair of individuals, and can be estimated accurately for high coverage data by comparing genotypes. For low coverage samples, $D_w$ is typically calculated by comparing randomly-sampled reads (or pseudo-haploid sequences) \cite{haak_massive_2015}. However, such an approach may result in loss of data, and hence we calculate $D_w$ by implicitly summing over all possible samplings:
\begin{align}\label{eq:x}
D_w &= \sum_s \nu_i(s) (1-\nu_j(s)) + (1-\nu_i(s)) \nu_j(s)
\end{align}
Here, $\nu_i(s)$ and $\nu_j(s)$ are the derived allele frequencies for individuals i and j respectively, at SNP index $s$ in a window.
Throughout, we will use bold-face notation to refer to the vector (or matrix) collecting  all the terms, e.g. $\BD = (D_1, D_2, \dots D_L)$. The KIN-HMM uses $\BD$ and $\BN$  to classify each window into three hidden states $Z_w \in$ ($0$, $1$, $2$), reflecting zero, one or two chromosomes shared IBD. Likewise, we classify each window into one of three possible ROH state $H_w \in$ ($0$, $1$, $2$), reflecting, zero, one or both individuals being in a run-of-homozygosity at this genomic location.

There are two additional parameters, $\pi$ and $\mathbf{A}$. The initial probabilities $\pi$ give the probabilities of being in each state $Z$ at the beginning of each chromosome, which we set to uniform for simplicity. The transition matrix $\mathbf{A}$ gives the probability of moving from state $i$ to state $j$, given by $a_{ij}$, and is fixed for each relatedness category (see section YYY). For compactness of notation, we group the fixed parameters: $\theta = (\BN, \mathbf{A}, \pi)$.

@@

\begin{align}\label{eq:1}
\log P(\BD,\BZ|\BH, \theta) &= \log P(\BD|\BZ,\BH, \theta) + \log P(\BZ |\theta) \nonumber\\
&= \sum_w \log P(D_w|Z_w,H_w, \theta) + \sum_w \log P(Z_w |Z_{w-1},\theta) + \log P(Z_0|\theta)
\end{align}

\subsubsection{Emission probability}
Using this set up, we can isolate the emissions $P(D_w | Z_w, H_w ,\theta)$ from equation \ref{eq:1} and optimize them alone. The number of shared sites $N_w$ is the only parameter in $\theta$ that affects the emissions.

Assuming sites are equally distributed and independent would result in the binomial likelihood:
$$P(D_w|Z_w, H_w, N_w) \sim \text{Binom}[D_w ; p(Z_w, H_w), N_w] \text{,}$$

where, $p$ is the expected proportion of differences we would expect for a particular IBD and ROH state. If the two individuals are unrelated in a particular window (i.e. $Z_w = 0$), then the expected proportion of pairwise differences depends solely on the effective population size, and we denote this proportion with $p_0$. Note that $p_0$ does not depend on the presence of ROH, since this does not change expected proportion of differences between unrelated individuals. If the two individuals share one or even both copies of the genome IBD, we would expect the proportion of differences to be reduced to $p_1 = \frac{3}{4} p_0$, and $p_2 = \frac{1}2 p_0$, respectively, since either one or two of the four possible comparisons will be between identical chromosomes \cite{kuhn_estimating_2018}. Thus, $p(Z_w=i, H_w=0) = p_i$.

The proportion of differences between unrelated individuals $p_0$ is an important parameter. We follow READ \cite{kuhn_estimating_2018} and estimate it as the median of differences for all possible pairs of individuals, which should work well if the majority of individuals in the samples are unrelated. We set $p_1$ to $(3/4)p_0$ and $p_2$ to $p_0/2$.

@@

\paragraph{Beta Binomial Model}
The Binomial Model laid out above assumes that sites are independent. However, due to genetic linkage, nearby sites will be correlated and we find that the data often has considerably higher variance than would be expected from a binomial model. We take this into account by adding an overdispersion parameter $\delta$ that allows for variation in the $p$ between windows. Note that, just like $p(Z_w,H_w)$, $\delta(Z_w,H_w)$ has effectively four unique parameters $\delta_i$ with $i$ $\in$ (0,1,2,3). This is due to the fact that several combinations of $Z_w$ and $H_w$ can lead to comparison between same number of unique chromosomes, and such states will have same mean as well as the same variance. This results in the beta-binomial likelihood, that can be fitted to the data (fig. S\ref{figS1:binom})).

\begin{align}\label{eq:3}
P(D_{w}|p_i,\delta_i,N_w) &\sim BB[D_w; N_w, p_i, \delta_i] \nonumber\\
&= \binom{N_w}{D_w}\frac{B(D_w+p_x \delta_{i}, N_w-D_w+ \delta_{i}(1-p_{i}))}{ B(p_{i}\delta_{i}, (1-p_{i})\delta_{i})}.
\end{align}

This parameterization of the beta distribution in terms of expected value $p$ and overdispersion $\delta$ is also called the Balding-Nichols-model \cite{balding_method_nodate}, and is distinct from the more common parametrization in terms of $\alpha$ and $\beta$. It is worth mentioning here that $D_w$ is estimated as a sum of probabilities (see section XX), while both $D_w$ and $N_w$ may undergo contamination correction, which causes these estimates to be non-integral. We calculate factorials required in Beta Binomial likelihood with Gamma function, which is continuous and hence does not require integral values.

\subsubsection{Estimation}

Emission probabilities are estimated with Expectation-Maximization (EM) algorithm. In each iteration of the algorithm, the Expectation step involves forward-backward algorithm to calculate $P(D, Z| H, \theta, \delta)$.

In maximization step we use these joint probabilities to optimize $\delta_i$ for each state i $\in$ (0,1,2,3), and update emission probabilities. Iterations are run, until the complete data log likelihood stops changing.

\paragraph{Initialization}
The value of $\delta_i$ is unknown to start with, and is set to a random value.

\paragraph{Expectation step}
In the $t$-th iteration, we calculate the posterior probability of each IBD state in each window $\gamma^{(t)}_{wi} = P(Z_w=i | \BD, \BH, \theta, \delta^{(t)})$ using the forward-backward algorithm, where $\delta^{(t)}$ is the current guess for delta.

\paragraph{Maximization step}
The only free parameters we estimate in the M-step are the overdispersion parameters $\delta_i$. We do this optimization using a cost function, which is the log-emission probability weighted by the posterior probabilities of the hidden states $\gamma_{wi}$ and the ROH state-probabilities $h_{w\omega}$.

\begin{align}\label{eq:4}
\mathcal{C} &= \mathbb{E}[\log P(D_w|Z_w, H_w, \theta, \delta^{(t-1)}]\nonumber\\
&= \sum_{w=1}^L \sum_{i=0}^2\sum^2_{\omega=0} \log P(D_{w}|Z_w, H_w, N_w, \delta^{(t-1)}, p)h_{w\omega}\gamma_{wi}
\end{align}

Using equation \ref{eq:2}, we simplify this by grouping all the terms that would result in the same $p$, i.e.

@@

So we can rewrite equation \ref{eq:4} as:
\begin{align}\label{eq:6}
\mathcal{C} &= \sum_{w=1}^L\sum_{i=0}^3 \log P(D_{w}|Z_w, H_w, N_w, \delta_i, p_i)g_{wi}
\end{align}

Since each term in equation \ref{eq:6}  depends on only one $\delta_i$ and not on different IBD and ROH states, we can optimize for them independently.

We realized that different cases of relatedness have different number of IBD states possible. For example, siblings may have all three IBD states present while a Parent-Child has only $Z_w=1$. This may cause some overfitting problem, when we apply a Parent-Child HMM to a dataset corresponding to siblings. We show a simple example of this problem in fig. \ref{figS4:bndsbeta}. We avoid this problem by constraining the parameter space of $\delta$. 
In particular, the variance of a beta distributed random variable $X$ is:
\begin{align}\label{eq:7}
    var(X) = \frac{p(1-p)}{\delta + 1}
\end{align}
We constrain $\delta$ such that  the variance is less than a threshold T:

\begin{equation}\label{eq:8}
    \frac{p(1-p)}{\delta + 1}  < T
\end{equation}

We choose T for each beta distribution such that the distributions do not overlap much.

>>>>

\title{KIN: Method to infer relatedness using low coverage ancient DNA}
\author{Divya Ratan Popli, Stéphane Peyrégne, Benjamin M. Peter}
\date{5 August 2021}
\linenumbers

@@

\noindent Knowledge of familial relationships is an important part of several research fields. Knowing genetic kinship of ancient specimens can lead to interesting insights into the culture and social hierarchy of ancient populations, and is also relevant for downstream genetic analyses. However, estimating relatedness from ancient DNA can be difficult since DNA sequences extracted from ancient remains may have low coverage, ascertainment bias, and contamination from modern human DNA. In addition, the population of interest may have long runs of homozygosity (ROH) due to recent inbreeding or/and small population size. This can affect the relatedness estimates. Here, we present a Hidden Markov Model (HMM) to estimate identity by descent (IBD) fragments and degree of relatedness up to 3rd degree, while additionally differentiating between siblings and parent-child with as low as 0.05x coverage DNA data. The output can be further refined with another HMM that we developed to estimate ROH proportions. We focus on the application of KIN on ancient human DNA, but it is applicable to other diploid species as well. On a machine with 8 cores, the generation of input files from 17 bam files takes -- minutes, and running KIN software takes -- minutes.
\end{abstract}

@@

Identifying related individuals is a common task in genetic studies. Relatedness is of direct interest in e.g. DNA forensics, where familial search can aid in solving criminal cases, and to identify unknown deceased persons \cite{murphy_law_2018,ram_genealogy_2018}. Genetic paternity tests have an important application in resolving family relation, e.g. in establishing relationship between a person applying for immigration and the claimed relatives \cite{egeland_beyond_2000}. It is also an essential pre-processing step in population genetics and association studies, where samples are typically assumed to be independent random draws from the population.

In ancient DNA studies, relatedness can be used to identify bones and teeth belonging to the same individual, and can provide an understanding of an ancient society's organization and hierarchy, social structures, and cultural aspects ~\cite{baca_ancient_2012,mittnik_KINship-based_2019,sikora_ancient_2017}. For animal and plant breeders and conservation biologists, reconstructing pedigrees and finding related individuals is important to avoid inbreeding and ensure diversity. ~\cite{habier_impact_2007,oliehoek_estimating_2006,kardos_measuring_2015}

\subsection{Approaches to estimate relatedness from high coverage data}

Commonly, pairs of related individuals are identified by looking for parts of the genome that are identical by descent (IBD), ie. inherited from a recent common ancestor. Due to the laws of Mendelian segregation, each parent, for example, will share exactly one set of chromosomes IBD with their offspring. A grandparent will, on average share a quarter of their genome with a grand-child due to recombination. Along the genomes of a pair of diploid individuals, there are three IBD states possible at any given position: either no chromosomes are IBD, only one chromosome is IBD, or both chromosomes are IBD. The genome-wide proportions of these states (usually referred to as $k_0$, $k_1$, $k_2$, so that $k_0+k_1+k_2=1$) can be used to infer the degree and nature of relatedness for a pair of individuals. For example, a pair of siblings are expected to have all three possible IBD states with probabilities (0.25,0.5,0.25) in the absence of inbreeding, as shown in Fig \ref{fig0:schematic}. These IBD probabilities can directly be used to categorize their relatedness as shown in table \ref{tab:Table 1}. One can also use these probabilities to estimate the coefficient of relatedness $r$, which is defined as the proportion of the genome that is IBD. In the absence of inbreeding, this would be calculated as $r= k_1/2 + k_2$.

However, it is not possible to directly observe IBD segments, and so most methods infer them by first identifying segments of the genome that are Identical by State (IBS) and using population allele frequencies to calculate the probability of IBD given IBS \cite{vai_KINship_2020}. One exception is SNPduo, a software that uses IBS directly from a pair of individuals to visualize and analyze their relatedness. \cite{roberson_visualization_2009}. Most methods, however, use IBD as a hidden state in inferring relatedness \cite{boehnke_accurate_1997,lynch_estimation_1999, albrechtsen_natural_2010, purcell_plink_2007}.

@@

These methods utilize population allele frequencies, phase information, recombination maps, or genotype calls to co-estimate IBD and the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,li_accurate_2014,thornton_estimating_2012}. There are many such methods available. PLINK, a popular tool to infer relatedness coefficient in diploid genotype data, estimates IBD probabilities from observed IBS, using the allele frequencies at each Single Nucleotide Polymorphism (SNP) in the analyzed samples \cite{purcell_plink_2007}. KING is another widely used software that allows relatedness inference with genotype data in homogeneous populations using population allele frequencies in the samples (KING homo), but can also work with structured populations with no good population allele frequencies using heterozygosity estimates from each sample (KING robust) \cite{manichaikul_robust_2010}. .

\begin{table}
\caption{\label{tab:Table 1}IBD sharing probabilities for different relations in absence of inbreeding}
\begin{tabular}{|c|c|c|c|}
    \hline
    Relatedness & $k_0$ & $k_1$ & $k_2$\\
    \hline
    Unrelated & 1 & 0 & 0\\
    \hline
    3rd Degree & 0.75 & 0.25 & 0\\
    \hline
    2nd Degree & 0.5 & 0.5 & 0\\
    \hline
    Siblings & 0.25 & 0.5 & 0.25\\
    \hline
    Parent-Child & 0 & 1 & 0\\
    \hline
    Identical/twins & 0 & 0 & 1\\
    \hline
\end{tabular}
\label{table1}
\end{table}

\subsection{Approaches that address problems with ancient DNA data}
One major issue with applying the above mentioned methods to ancient DNA data is that the sequence coverage is typically low, making it difficult to get genotype calls.
Several methods surmount this problem by using genotype likelihoods \cite{lipatov_maximum_2015,korneliussen_ngsrelate_2015}. In this way, it is possible to account for the uncertainty in genotype calls by summing over all possible genotypes weighted by their genotype likelihoods. However, these approaches typically still require at least 2x coverage, since genotype likelihoods may be unreliable at lower coverages. Ancient DNA analyses often face additional challenges such as contamination from present-day DNA \cite{peyregne_authentict_2020}, absence of a good reference panel, and ascertainment bias in the case of data generated from DNA capture. In cases where the allele frequency or a reference panel for the target population is not available, it is possible to use allele frequency from a modern population from the same geographic location, or to try using population allele frequencies from multiple potentially close populations. However, incorrect assumptions may cause the individuals in the target population to look more similar or different to each other \cite{amorim_understanding_2018}. Several methods have been proposed to estimate relatedness without a reference panel, but many of these require either $>4$x coverage \cite{waples_allele_2019}, or a large sample size to get an estimate of population allele frequencies from the samples \cite{theunert_joint_2017}. In this study, we show that contamination can cause some of the individuals in ancient populations to look more unrelated than they are. This happens because the DNA sequence of an ancient individual typically exhibits more differences to a contaminant sequence than that of another ancient individual from the same population. It is worth mentioning that different DNA libraries may have different levels of contamination, even when they have been sampled from the same specimen. Hence, contamination can reduce the power to detect related pairs, especially when the target population is quite diverged from the contaminating population(s). In addition, the genomes of target individuals may have long runs of homozygosity (ROH) due to a small population size, or recent inbreeding. Long ROH cause related individuals to seem genetically closer to each other, but do not affect the genetic distance between unrelated individuals. This effect may cause an increase in falsely inferred relationships. Moreover, ancient DNA is commonly captured with a SNP array to enrich for sequences from the target species. The way that these SNPs are defined (e.g. selecting SNPs variable in certain individuals) becomes relevant, because methods to infer relatedness based on the fraction of sites in different IBS states are highly sensitive to ascertainment bias, and may give different results when applied to data with different ascertainment schemes \cite{waples_allele_2019}.

@@

\section{New Approaches}\label{new_approaches}

To infer the case of relatedness, KIN fits one HMM for each pair of individual and for each relatedness case. The KIN-HMM infers IBD sharing between a pair of low-coverage individuals, optionally taking ROH tracts and contamination estimates in each individual into account. The best-fit model is then assigned as the inferred relatedness. If the locations of ROH tracts are unknown, we provide another HMM (ROH-HMM) to coarsely estimate the location of ROH for samples with sufficient coverage ($\geq 0.1x$). Our method is available on \url{https://github.com/DivyaratanPopli/KIN_snakemake} along with a \textt{snakemake} \cite{koster_snakemakescalable_2012} pipeline to generate the input files for the models directly from bam files.

\subsection{Model description}\label{method_overview} 
To infer the IBD-state across the genome, we subdivide the genomes of a pair of individuals into $L$ large genomic windows (typically of size 10Mb). The proportion of pairwise differences in those windows will differ based on the IBD-state, and the relatedness case will determine how often we expect to switch between different IBD-states.  The cases we consider are:  Unrelated, $5^{th}$ Degree, $4^{th}$ Degree, $3^{th}$ Degree, Grandparent-Grandchild, Avuncular, Half-siblings, Parent-Child, Siblings, and Identical individuals. The details of the likelihood computation are given in section \ref{ll}. The likelihood computations require an emission (section \ref{B}) and transition probability.  We estimate the transition matrix for each relatedness case from simulations (see section \ref{simulat}), and hence the transition probability for a HMM is fixed. To calculate the emission probabilities, we need to estimate a parameter $\delta$ that describes the variance in the data given the IBD state and the ROH state. The estimation of this parameter is essential to our inference of relatedness, since the variance in the data can change with the number of overlapping sites. So, our algorithm optimizes $\delta$, which is used in the calculation of emission probabilities, and we use these emission probabilities, along with the transition probabilities, to estimate hidden IBD states and to calculate the log likelihood for the HMM. We can then compare the log likelihood from each model to find the best fit along with a measure of uncertainty. We output the IBD states corresponding to the most likely model. 
Hidden IBD state of each window ($Z_w$) is estimated using standard Viterbi algorithm, and $\BZ$ corresponding to the most likely model are returned.

\subsection{Input of KIN-HMM}\label{hmm_input}
The inputs of our algorithm are i) the number of overlapping sites for the $w$-th window $N_w$ for which both samples have at least one read available, ii) the expected number of pairwise differences $D_w$ at these sites, and iii) the probability of ROH in windows, by default obtained from ROH-HMM described in section \ref{roh}.

$D_w$ is the number of sites that are different between a pair of individuals, and can be estimated accurately for high coverage data by comparing genotypes. For low-coverage samples, $D_w$ is often calculated by comparing randomly-sampled reads ("pseudo-haploid" sequences) \cite{haak_massive_2015}. However, such an approach may result in loss of data, and hence we calculate $D_w$ by implicitly summing over all possible samplings:
\begin{align}\label{eq:x}
D_w &= \sum_s \nu_i(s) (1-\nu_j(s)) + (1-\nu_i(s)) \nu_j(s)
\end{align}
Here, $\nu_i(s)$ and $\nu_j(s)$ are the proportions of reads carrying the derived allele at SNP index $s$ for individuals i and j respectively.
Throughout, we will use bold-face notation to refer to the vector (or matrix) collecting  all the terms, e.g. $\BD = (D_1, D_2, \dots D_L)$.

\subsection{Log likelihood of the KIN-HMM}\label{ll}
The KIN-HMM uses $\BD$ and $\BN$ to classify each window into three hidden states $Z_w \in$ ($0$, $1$, $2$), reflecting zero, one or two chromosomes IBD. To take ROH into account, we define the variable $H_w \in$ ($0$, $1$, $2$) that designates that zero, one or both individuals are homozygous in window $w$. Since $H_w$ is unobserved, in practice we use the estimates from ROH-HMM $h_{wj} = P(H_w = j)$.

There are three additional model parameters, $\pi, \mathbf{A}$ and $\delta$. The initial probabilities $\pi$ give the probabilities of being in each state $Z$ at the beginning of each chromosome, which we set to uniform for simplicity. The transition matrix $\mathbf{A}$ gives the probability of moving from state $i$ to state $j$, given by $a_{ij}$, and is fixed for each relatedness case. The overdispersion-parameter $\delta$ takes into account that SNPs in each window vary in their allele frequency (see next setion).  For compactness of notation, we group the fixed parameters: $\theta = (\BN, \mathbf{A}, \pi)$.

@@

\begin{align}\label{eq:1}
\log P(\BD,\BZ|\BH, \theta, \delta) &= \log P(\BD|\BZ,\BH, \theta, \delta) + \log P(\BZ |\theta) \nonumber\\
&= \sum_w \log P(D_w|Z_w,H_w, \theta, \delta) + \sum_w \log P(Z_w |Z_{w-1},\theta) + \log P(Z_0|\theta)
\end{align}

\subsection{Emission probability}\label{B}

Using this set up, we can isolate the emissions $P(D_w | Z_w, H_w ,\theta, \delta)$ from equation \ref{eq:1} and optimize them alone. The number of shared sites $N_w$ is the only parameter in $\theta$ that affects the emissions.

Assuming the sites were equally distributed and independent would result in the binomial likelihood:
$$P(D_w|Z_w, H_w, N_w) \sim \text{Binom}[D_w ; p(Z_w, H_w), N_w] \text{,}$$

where, $p$ is the proportion of differences expected for a particular IBD and ROH state. If the two individuals are unrelated in a particular window (i.e. $Z_w = 0$), then the expected proportion of pairwise differences depends solely on the effective population size, and we denote this proportion with $p_0$. Note that $p_0$ does not depend on the presence of ROH, since this does not change expected proportion of differences between unrelated individuals. If the two individuals share one or even both copies of the genome IBD, we would expect the proportion of differences to be reduced to $p_1 = \frac{3}{4} p_0$, and $p_2 = \frac{1}2 p_0$, respectively, since either one or two of the four possible comparisons will be between identical chromosomes \cite{kuhn_estimating_2018}. Thus, $p(Z_w=i, H_w=0) = p_i$.

The proportion of differences between unrelated individuals $p_0$ is an important parameter. We follow READ \cite{kuhn_estimating_2018} and estimate $p_0$ as the median of differences for all possible pairs of individuals, which should work well if the majority of individuals in the samples are unrelated. We set $p_1$ to $(3/4)p_0$ and $p_2$ to $p_0/2$.

@@

\paragraph{Beta Binomial Model}
The Binomial Model laid out above assumes that sites are independent. However, due to genetic linkage, nearby sites will be correlated and we find that the data often has considerably higher variance than would be expected from a binomial model. We take this into account by adding an overdispersion parameter $\delta$ that allows for variation in the $p$ between windows. Note that, just like $p(Z_w,H_w)$ the overdispersion $\delta(Z_w,H_w)$ depends on the number of chromosomes compared, there are only four parameters $\delta_i$ with $i$ $\in$ (0,1,2,3) analogous to the $p_i$. Empirically, we find that the beta-binomial likelihood provides a good fit to the data (fig. S\ref{figS1:binom})).

Taken together, our emission probabilities are 
\begin{align}\label{eq:3}
P(D_{w}|Z_w,H_w,N_w, \delta) = P(D_{w}|p_i,\delta_i,N_w) &\sim BB[D_w; p_i, \delta_i, N_w] \nonumber\\
&= \binom{N_w}{D_w}\frac{B(D_w+p_i \delta_{i}, N_w-D_w+ \delta_{i}(1-p_{i}))}{ B(p_{i}\delta_{i}, (1-p_{i})\delta_{i})},
\end{align}
where $i$ is fully determined by the combination of $Z_w$ and $H_w$ (see eq 3).

This parameterization of the beta distribution in terms of expected value $p$ and overdispersion $\delta$ is also called the Balding-Nichols-model \cite{balding_method_nodate}, and is distinct from the more common parametrization in terms of $\alpha$ and $\beta$. We use this equation as well if some preprocessing steps result in non-integer $D_w$ and $N_w$, in which case we approximate the binomial coefficient using Gamma functions.

\subsection{Estimation of $\delta$}\label{delta}
We estimate the $\delta$-parameters using an Expectation-Maximization (EM) algorithm (REF). In an EM-algorithm, we iterate the Expecation and Maximiation Steps until the algorithm converges.

\paragraph{Initialization}
The value of $\delta_i$ is unknown to start with, and we set it to a random value.

\paragraph{Expectation step}
In the $t$-th iteration, we calculate the posterior probability of each IBD state in each window $\gamma^{(t)}_{wi} = P(Z_w=i | D_w, H_w, \theta_i, \delta_i^{(t)})$ using the forward-backward algorithm, where $\delta_i^{(t)}$ is the current guess for $\delta$ for a given IBD state.

\paragraph{Maximization step}
The only free parameters we estimate in the M-step are the overdispersion parameters $\delta_i$. We do this optimization using a cost function, which is the log-emission probability weighted by the posterior probabilities of the hidden states $\gamma_{wi}$ and optionally the ROH state-probabilities $h_{w\omega}$ obtained from the ROH-HMM.

\begin{align}\label{eq:4}
\mathcal{C} &= \mathbb{E}[\log P(D_w|Z_w, H_w, \theta, \delta^{(t-1)}]\nonumber\\
&= \sum_{w=1}^L \sum_{i=0}^2\sum^2_{\omega=0} \log P(D_{w}|Z_w, H_w, N_w, \delta_{i \omega}^{(t-1)}, p_{i \omega})h_{w\omega}\gamma_{wi}
\end{align}

Using equation \ref{eq:2}, we simplify this by grouping all the terms that would result in the same $p_i$ and $\delta_i$, i.e.

@@

So we can rewrite equation \ref{eq:4} as:
\begin{align}\label{eq:6}
\mathcal{C} &= \sum_{i=0}^3 \sum_{w=1}^L \log P(D_{w}|Z_w, H_w, N_w, \delta_i, p_i)g_{wi}
\end{align}

The cost function \ref{eq:6} has one independent term for each $i$ and so we can separate them and estimate each $\delta_i$ independently using the XXX algorithm implemented in \texttt{scipy.optimize} (REF scipy).

%\paragraph{Constraints on $\delta$ optimization}\label{delta_constraint}

Unconstrained optimization of the $\delta_i$ could result in some confounding of cases, We realized that different cases of relatedness have different number of IBD states possible. For example, siblings may have all three IBD states present while a Parent-Child has only $Z_w=1$. However, the parent-child model could fit data generated under the sibling model by assigning it a very high $\delta_1$, which would reduce performance  \ref{figS4:bndsbeta}. We avoid this problem by constraining the $\delta$ such that the beta distributions for the different cases overlap by at most one standard deviation.

\subsection{Model comparison}\label{model_comp}
To infer the most probable relatedness case, we run our model on all cases ( section \ref{method_overview}) and compare the resulting likelihoods. We output the relatedness corresponding to the maximum likelihood model, and the confidence as given by the log likelihood ratio between the two highest likelihood models. The log likelihood-ratio is commonly used in likelihood-ratio tests to compare nested models. However, in our cases, models are not nested, and so standard likelihood-ratio theory does not apply. For this reason, we use the log likelihood ratio between the two best models as a statistic to assess the confidence in our classifications, and use simulations to obtain critical values (Fig \ref{fig3:cutoff}).

\paragraph{Grouping of cases}
Particularly for low-quality data, we may not be able to distinguish all cases. Thus, we group $4^{th}$ and $5^{th}$ degree relatives together with unrelated.  Similarly, we group  half-siblings, avuncular and grandparent-grandchild to  $2^{nd}$-degree relatives in the final results. We report the final pairwise classification in the following categories: Unrelated, Third Degree, Second Degree, Parent-Child, Siblings, Identical individuals.
